---
title: "Bessel’s Correction"
author: "My Name"
output: html_document
---

_**In the context of sampling, Bessel's Correction improves the estimate of standard deviation: specifically, while the sample standard deviation is a biased estimate of the population standard deviation, the bias is smaller with Bessel's Correction.**_

```{r load, message=F, echo=F}
require(ggplot2)
require(dplyr)
require(scales)
require(reshape2)
require(lubridate)
require(ggthemes)
require(gridExtra)
require(lubridate)
data(diamonds)
```

For data, we will use the _diamonds_ data set in the R-Package ggplot2, which contains data from `r nrow(diamonds)` round cut diamonds. Here are the first 6 rows of this data set:
```{r head, echo=F}
head(diamonds)
```
## Describing the distribution of the "price" variable

<span style="color:red">Answer this question: what is the meaning of a distribution of a variable, and how does it relate to price?</span> 

#### Type of variable chosen

<span style="color:red">Explain what a quantitative variable, and why it was important to make such a choice in a report about standard deviation.  Explain how the concepts of numerical and quantitative variables are different, though related.</span> 

#### Histogram of diamonds price.

<span style="color:red">What is a histogram?  Explain graph below.</span> 

```{r histdia, echo=F}
ggplot(diamonds, aes(x = price)) + 
  geom_histogram(color = "black", fill = "DarkOrange", binwidth = 500) + 
  scale_x_continuous(labels = dollar, breaks = seq(0, 20000, 1000)) + 
  theme(axis.text.x = element_text(angle = 90)) + 
  xlab("Price") + ylab("Count")
```

#### Violin plot

<span style="color:red">Explain the relationship between a histogram and a violin plot.</span> 

```{r violdia, echo = F}
ggplot(diamonds, aes(x=0, y=price)) + geom_violin(fill = "DarkOrange") + xlab("") + theme(axis.text.x=element_blank())
```

#### Numerical Summaries

R has a function that returns numerical summaries of data.  For example:

```{r summary}
summary(diamonds$price)
```

<span style="color:red">Describe what each of these numbers means.</span> 

#### Modified Box Plots
<span style="color:red">Describe the relationship of the numbers above to the modified box plot, here drawn inside the violin plot.  Explain the difference between a boxplot and a modified box plot. </span> 

```{r boxdia, echo = F}
ggplot(diamonds, aes(x=0, y=price)) + geom_violin() + geom_boxplot(width=0.1, fill = "DarkOrange") + xlab("") + theme(axis.text.x=element_blank())
```

#### Adding the mean to the plot
<span style="color:red">Add one sentence to indicate where the mean is on this plot.</span> 

```{r meandia, echo = F}
ggplot(diamonds, aes(x=0, y=price)) + geom_violin() + geom_boxplot(width=0.1) + xlab("") + theme(axis.text.x=element_blank()) + stat_summary(fun.y=mean, geom="point", size=2, color="red")
```

#### Standard Deviation: Formulas

<span style="color:red">Explain the formulas below, say which uses Bessel's correction.</span> 

$$s = \sqrt{\frac{1}{n-1}\sum\left(x_i - \bar x\right)^2}$$

$$s_n = \sqrt{\frac{1}{n}\sum\left(x_i - \bar x\right)^2}$$

#### Standard Deviation of Diamonds Price
We compute the standard deviation (with Bessel's correction) of the price variable:

```{r sd}
sd(diamonds$price)
```

How about without Bessel's correction?  Well, R doesn't seem to have this function, but we can add it:

```{r sdn}
sdn <- function(x) {
  return(sqrt(mean((x - mean(x))^2)))
}
sdn(diamonds$price)
```

## Sampling

The statement that began this document asserted that Bessel's correction is important in the context of _sampling_. <span style="color:red">Explain sampling here, explain the difference between the sample standard deviation and the population standard deviation, and say which we used above.  Explain the differences between a population, and a sample, a parameter and a statistic.  G</span>

sampling distributions for statistics, sampling error, sampling bias, biased estimators and unbiased estimators.
