---
title: "GGPlot2 Examples"
author: "Sean Carver"
date: "11/1/2017"
output: html_document
---

The first step is to load data. Make sure you have installed the packages "ggplot2" and "car".
```{r loaddata}
library(ggplot2)
data(Salaries, package="car")
```

After loading data, the next step is to create the base layer with the function ggplot().  You typically pass two arguments to ggplot: the data and the asethetics (i.e. what features in the data generate what features in the plot).

The data is: Salaries.  The asethetics for this first plot are: x goes with rank, and y goes with salary.  We are going to do a boxplot with academic rank on the x-axis and salaries on the y-axis.

We are going to call the layers gg as we build them.  Each time we are going to plot the layer by calling its name.

```{r plotsalaryvsrank}
gg <- ggplot(Salaries, aes(x=rank, y=salary))
gg
```

Now we are going to add a geometic object, but we are just going to plot it: we are not going to save it until later.
```{r addboxplot}
gg + geom_boxplot()
```

Let's try another color.
```{r colorboxplot}
gg + geom_boxplot(fill="cornflowerblue")
```

Let's add the notches.  If notches don't overlap, we have evidence that the medians are truly different across the population, even if our data just come from a random sample.

```{r addnotch}
gg + geom_boxplot(fill="cornflowerblue", notch=TRUE)
```

I'm happy with this plot.  Let's save it so we can add a third layer.  This won't create a graph, unless you add the command "gg"

```{r saveboxplot}
gg <- gg + geom_boxplot(fill="cornflowerblue", notch=TRUE)
```

Now we add a scatterplot (without saving):

```{r addscatter}
gg + geom_point()
```

Will be easier to see if we jitter the dots, so that they are not all in a single line.

```{r addjitter}
gg + geom_point(position="jitter")
```

Black dots are too harsh.  Let's make them blue.

```{r makedotsblue}
gg + geom_point(position="jitter", color="blue")
```

Now let's make them slightly transparent.  That is done with "alpha" 0 for completely transparent 1 for complete opaque.  Let's go with 0.5

```{r makedotstransparent}
gg + geom_point(position="jitter", color="blue", alpha=0.5)
```

Now save it so we can add a forth layer.

```{r savethirdlayer}
gg <- gg + geom_point(position="jitter", color="blue", alpha=0.5)
```

Now let's add a rug plot

```{r addrug}
gg + geom_rug()
```

We could save this layer to the variable gg, but there is no need, because we have already plotted it and we don't need it again, for anything else.

So instead let's start over with a new base layer.  This will be a density plot.  We are going to plot the density on the y axis and the salary on the x axis.  There will be there density plots of different colors (various "fill"'s).  We don't need to define an aesthetics for y because it is implied by density.  Remember this is a new base layer (no sum).

```{r densitybaselayer}
gg <- ggplot(data=Salaries, aes(x=salary, fill=rank))
gg
```

Now add density object:

```{r adddensity}
gg + geom_density()
```

These would be better transparent:

```{r makedensitiestrans}
gg + geom_density(alpha=0.3)
```

Let's make a new plot.  We are going to do a map of murder rates in the US in 1973, the data are in USArrests.  The rows are labeled by state, however we need a column of states, not labels so we can add one with:
```{r states}
data(USArrests)
crimes <- data.frame(state=tolower(rownames(USArrests)), USArrests)
```

Now let's do a new base layer.

```{r statesbaselayer}
gg <- ggplot(crimes, aes(map_id=state, fill=Murder))
gg
```

Let's add a layer, first save, then plot.

```{r addgeomstate}
gg <- gg + geom_map(map=map_data("state"))
gg
```

This still doesn't do anything except plot a legend.  We need to associate x and y.

```{r makemap}
gg <- gg + expand_limits(x=map_data("state")$long, y=map_data("state")$lat)
gg
```

I am going to make this better by adding labels:

```{r addlabels}
gg + labs(title="Murder rates per 100,000 in 1973", x="logitude", y="latitude")
```

