---
title: "Quiz 2 Prep"
author: "STAT 202"
output: html_document
---

Here are some of the questions I might ask.

## Sampling Distributions

Look at the following graphs.

```{r load, message=F, echo=F, cache=FALSE}
require(ggplot2)
require(scales)
require(reshape2)
require(lubridate)
data(diamonds)
```

```{r sdn, echo=F}
sdn <- function(x) {
  return(sqrt(mean((x - mean(x))^2)))
}
```

```{r samplesizeplot, echo=F}
diamonds.price.sample.of.n <- function(sample.size) {
  return(sample(diamonds$price, sample.size))
}

m.statistics.of.n <- function(statistic,number.of.samples,sample.size) {
  stats <- c()
  for (k in 1:number.of.samples) {
    stats <- c(stats, statistic(diamonds.price.sample.of.n(sample.size)))
  }
  return(stats)
}

set.seed(2)
m <- 1000
stats <- cbind()
labels <- cbind()
levels <- c(1,4,16,64,256,1024)
for (i in c(1,4,16,64,256,1024)) {
  new.stats <- cbind(m.statistics.of.n(statistic=mean, number.of.samples=m, sample.size=i))
  new.labels <- cbind(rep(as.character(i), m))
  stats <- rbind(stats, new.stats)
  labels <- rbind(labels, new.labels)
}
df <- data.frame(stats,labels)
df$labels <- factor(df$labels,levels=as.character(levels))
ggplot(df, aes(x=labels, y=stats)) + geom_violin() + geom_boxplot(width=0.1) + xlab("Sample Size") + ylab("Sample Mean") + stat_summary(fun.y=mean, geom="point", size=2, color="red") + geom_hline(yintercept=mean(diamonds$price))
```

```{r samplesd, echo=F, warning=F}
set.seed(2)
m <- 1000
stats <- cbind()
labels <- cbind()
levels <- c(1,4,16,64,256,1024)
for (i in c(1,4,16,64,256,1024)) {
  new.stats <- cbind(m.statistics.of.n(statistic=sd, number.of.samples=m, sample.size=i))
  new.labels <- cbind(rep(as.character(i), m))
  stats <- rbind(stats, new.stats)
  labels <- rbind(labels, new.labels)
}
df <- data.frame(stats,labels)
df$labels <- factor(df$labels,levels=as.character(levels))
ggplot(df, aes(x=labels, y=stats)) + geom_violin() + geom_boxplot(width=0.1) + xlab("Sample Size") + ylab("Sample Standard Deviation") + stat_summary(fun.y=mean, geom="point", size=2, color="red") + geom_hline(yintercept=sd(diamonds$price)) + scale_y_continuous(limits = c(0, 10000))
```

```{r samplesdn, echo=F, warning=F}
set.seed(2)
m <- 1000
stats <- cbind()
labels <- cbind()
levels <- c(1,4,16,64,256,1024)
for (i in c(1,4,16,64,256,1024)) {
  if (i==1) {
    new.stats <- cbind(rep(NA, m))
  } else {
    new.stats <- cbind(m.statistics.of.n(statistic=sdn, number.of.samples=m, sample.size=i))
  }
  new.labels <- cbind(rep(as.character(i), m))
  stats <- rbind(stats, new.stats)
  labels <- rbind(labels, new.labels)
}
df <- data.frame(stats,labels)
df$labels <- factor(df$labels,levels=as.character(levels))
ggplot(df, aes(x=labels, y=stats)) + geom_violin() + geom_boxplot(width=0.1) + xlab("Sample Size") + ylab("Sample Standard Deviation w/o Bessel's Correction") + stat_summary(fun.y=mean, geom="point", size=2, color="red") + geom_hline(yintercept=sdn(diamonds$price)) + scale_y_continuous(limits = c(0, 10000))
```

1. I will likely ask questions that have to do with interpreting these graphs.  For example, what is happening in the progression of sample size starting at 1 or 4 to 1024?  What difference in the properties of the estimator does the differences in the violin/box plot show?  What does this graph imply in terms of the quality of the estimates, and what are you trying to estimate?  Which graphs and which violin/boxes show approximately unbiased estimates?  Which show biased estimates?  What is the difference between sampling error and sampling bias?

2. What is a distribution of a variable and what the sampling distribution of a statistic?

3. Given an example of a random phenomenon: a game of chance, 1 person chosen at random from our class, 3 people chosen at random from our class, what is the sample space and what is an example of a random variable on that sample space?